Supplementary Material\\ \textit{Cross Validation Approaches for Penalized Cox Regression}

### 2019-05-22 main.tex
Edits to Section 3.2

The standard cross-validation in the partial likelihood setting is both straightforward and enjoys very good predictive accuracy, as seen in Section~\ref{Sec:accuracy}. However there are two main stability drawbacks to its widespread use. The following simulation studies illustrated those issues faced by the standard approach.

The first stability issue with the standard approach is that the number of uncensored observations in the test set may be insufficient to construct a well-defined partial likelihood, leaving the overall estimate of cross-validation error undefined as well. When computing the partial likelihood, only individuals whose events are observed will contribute to the likelihood. If there is no event observed in a particular fold, then the partial likelihood for that fold is undefined.

To illustrate how common this problem is, we conducted a simulation study. We generated 1000 independent data sets of $n = 100$ observations with fixed censoring rate. For each data set, we conducted a 10-fold cross-validation and recorded whether or not there was at least one event in each fold (i.e., whether the partial likelihood was well-defined for all folds). We repeated the same mechanism for scenarios with different censoring rates, ranging from $40\%$ to $80\%$. As illustrated in the left panel in Figure~\ref{Fig:stability}, at $n=100$ the standard approach can be problematic even at modest censoring rates. For example, standard CV cannot be used with over 20\% of data sets when $70\%$ of all observations are censored.

\begin{figure}[h]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_3.png}
  \caption{\label{Fig:stability} Simulations illustrating the stability issues with the standard approach. The figure on the left is generated with $n = 100$ observations. The horizontal axis is the censoring rate. The vertical axis is the proportion of undefined partial likelihood when the fold assignment is not balanced. The figure on the right is generated with $n = 120$. The fold assignment is balanced in this simulation so that the number of events in each fold is the same. The horizontal axis is the number of events per fold. The vertical axis is the standard deviation of the squared error loss. All four methods were compared in this simulation. The standard approach would have much greater variability when there are only few events in the fold.}
\end{figure}

One way could be used to get around the undefined partial likelihood for the standard approach is to balance the censoring status across all folds when making fold assignments. That is to make sure the number of events observed is about the same across all folds. Unless the number of the folds exceeds the number of observed events, such as the leave-one-out cross-validation, this approach will guarantee the partial likelihood in each fold is well-defined for the standard approach, thus the overall cross-validated error is well-defined.

Even if the partial likelihood could be well-defined with balanced fold assignment, there is another drawback of the standard approach: it tends to have more uncertainty in its selection. The predictive performance of the model selected by this approach has greaterr variation than the other approaches, especially when there are only a few events available to construct the partial likelihood.

The following simulation study was designed to illustrate this issue. We generated data sets with $n = 120$ observations. The censoring rate is fixed exactly at $50\%$, thus there are 60 events observed among all observations. For each data set, we conducted the balanced fold assignment with number of folds set to be 10, 12, 15, 20, 30 and 60. When the number of folds was set to be 10, there were exactly 6 events in each fold; when the number of folds was set to be 60, there was exactly one event in each fold. After each fold assignment, we computed the cross-validated error using all four approaches described in Section~\ref{Sec:accuracy}. We recorded the model that minimizes the cross-validated error for each approach and computed its squared error loss. We replicated the whole process for $N = 200$ time. In the end, we computed the standard deviation of the squared error loss for each approach.

The results of this simulation are illustrated in the right panel in Figure~\ref{Fig:stability}. While all other three approaches are not affected by the number of folds and number of events in each fold, the variability of the model performance has drastically changed for the standard approach. When there are sufficient number of events in each fold, the standard approach perform pretty close to the other three approaches. However, when there is only one or two events in each fold, the standard approach has much more variability in its predictive performance. Even if we can still use the standard approach when there are few number of events, there is a bigger uncertainty in terms of how well the predictive performance would be.

The 60-fold scenario in the last simulation study was also a corner example where there is only one event and one censored observation in each fold. In this scenario, it is common to run into cases where the uncensored event happens after the other event is censored. In this situation, the risk set at the time right before the event occurred would only contain this observation itself. Thus its contribution to the log partial likelihood would be 0, which is well-defined but not informative. This situation can happen regardless of the fold size and censoring rate. But it would matter more for the standard approach when there are fewer events in the fold.

Based on the two simulations illustrated in this section, we would make the following recommendations: i) always balance the censoring status when conducting fold assignments for the standard approach, ii) avoid using the standard approach in conjuction with high number of folds, iii) avoid using the standard approach when there are few number of events observed.

>>>>

Given that basic cross-validation is both straightforward and demonstrates strong predictive accuracy, as seen in Section~\ref{Sec:accuracy}, one might wonder why it is not more widely used for Cox regression.  The reason is that it suffers from several drawbacks with respect to numerical stability.

First, the number of uncensored observations in the test set may be insufficient to construct a well-defined partial likelihood for standard cross-validation, leaving the overall estimate of cross-validation error undefined as well. When computing the partial likelihood, only individuals whose events are observed will contribute to the likelihood. If there are no events observed in a particular fold, then the partial likelihood for that fold is undefined.

To illustrate how common this problem is, we generated 1000 independent data sets of $n = 100$ observations with a fixed censoring rate. For each data set, we conducted 10-fold cross-validation and recorded whether or not there was at least one event in each fold (i.e., whether the partial likelihood was well-defined for all folds). We repeated the same mechanism for scenarios with different censoring rates, ranging from $40\%$ to $80\%$. As illustrated in the left panel in Figure~\ref{Fig:stability}, at $n=100$ the standard approach can be problematic even at moderate censoring rates. For example, standard CV cannot be used with over 20\% of data sets when $70\%$ of observations are censored.

\begin{figure}[h]
  \centering
  \includegraphics[height= 7cm ]{./manuscript_figure/figure_3.png}
  \caption{\label{Fig:stability} Simulations illustrating the stability issues with the standard approach. For the figure on the left, $n = 100$, fold assignments are unbalanced, and the vertical axis is the proportion of undefined partial likelihoods. For the figure on the right, $n = 120$, fold assignments are balanced so that the number of events in each fold is the same, and the vertical axis is the standard deviation of the squared error loss. Basic cross-validation is much more variability than other approaches when there are only few events in each fold.}
\end{figure}

One approach for alleviating this issue is to balance the censoring status across all folds when making fold assignments. That is, to ensure the number of events observed is the same (or within 1) across all folds. Unless the number of the folds exceeds the number of observed events, such as in leave-one-out cross-validation, this approach will guarantee the partial likelihood in each fold is well-defined for the standard approach, thus the overall cross-validated error is well-defined.

In this case, however, even though the partial likelihood is well-defined, basic cross-validation suffers from extreme variability.  To illustrate, we generated data sets with $n = 120$ observations and a fixed censoring rate of $50\%$, so that there are 60 observed events. For each data set, we carried out cross-validation using balanced fold assignments with the number of folds set to be 10, 12, 15, 20, 30 and 60. When the number of folds was set to be 10, there were exactly 6 events in each fold; when the number of folds was set to be 60, there was exactly one event in each fold, and so on. After each fold assignment, we computed the cross-validated error using all four approaches described in Section~\ref{Sec:accuracy}. We recorded the model that minimizes the cross-validation error for each approach and computed its squared error loss. We replicated the whole process for $N = 200$ time. In the end, we computed the standard deviation of the squared error loss for each approach. \note{Does this phenomenon show up in the MSE too?}

The results of this simulation are illustrated in the right panel of Figure~\ref{Fig:stability}. While the other three approaches are not affected by the number of folds and number of events in each fold, the variability of basic cross-validation increases dramatically as the number of events per fold decreases. When there are sufficient number of events in each fold, the standard approach is similar to the other three approaches. However, when there are only one or two events in each fold, the standard approach can be up to 10 times more variable than other methods.  Thus, although the approach can be used to calculate CVE, there is considerable uncertainty surrounding its choice of the best model.

To shed some light on why the variability of basic cross-validation explodes in the 60-fold scenario, note that in this case, there is exactly one event and one censored observation in each fold. In many folds, the uncensored event happens after the other observation has been censored. In this situation, the risk set at the time immediately prior to the event contains only a single observation. Its contribution to the log partial likelihood is therefore 0, which is well-defined but not informative. This situation can happen regardless of the fold size and censoring rate, but it happens far more often as the number of events per fold decreases.

\note{Should we have this paragraph?  Seems to contradict our conclusion, in which our recommendation is to use the linear predictor approach.}  Based on the two simulations illustrated in this section, we would make the following recommendations: i) always balance the censoring status when conducting fold assignments for the standard approach, ii) avoid using the standard approach in conjuction with high number of folds, iii) avoid using the standard approach when there are few number of events observed.
